---
title: "Continuous Treatment in Propensity Score Analysis"
author: "Megha Joshi"
date: 2019-10-19
categories: ["propensity score analysis", "causal inference", "continuous treatment"]
bibliography: references.bib
csl: apa.csl
draft: false
---

In my qualifying exam, in the written part, I was asked about how to analyze the effect of continuous, not binary, treatment using propensity score analysis. I skipped it for the written but I spent a few days looking up how to analyze this in case I would be asked during my oral examination. Sadly, no one asked me even when I asked them to, so here is a blog detailing my explorations.

# Binary Treatment

For a review of propensity score analysis with binary treatment, please see [@stuart2010matching]. Below let $e(X)$ denote propensity scores, $D$ denote a binary treatment, and $X$ denote all the observed confounders. In the case of binary treatment, propensity scores represent the probability of receiving treatment given the covariates:

$$e(X) = P(D = 1|X)$$

We estimate the scores using logistic regression or machine learning techniques like generalized boosted models.

# Extension to Continuous Treatment

In binary treatment context, we assume that the potential outcomes ($Y(1)$ and $Y(0)$) are independent of treatment given $X$:

$$Y(1), Y(0) \perp\!\!\!\perp D |X$$

and by extension are independent given the propensity scores:

$$Y(0), Y(1) \perp\!\!\!\perp D|e(X)$$

@hirano_imbens_2004 introduced the assumption of weak unconfoundedness in the context of continuous treatment. They stated: "we do not require joint independence of all potential outcomes. Instead, we require conditional independence to hold for each value of the treatment." Below, let $T$ denote a continuous treatment variable. The potential outcome when $T = t$ is unrelated to the treatment given the set of covariates:

$$Y(t) \perp\!\!\!\perp T |X$$

To calculate the propensity scores, in the case of continuous treatment, we cannot find the probability that continuous treatment ($T$) equals a given value $t$. The likelihood of continuous variables taking on a given value is zero. For continuous treatment variable, we find the conditional density, the probability that $T$ is infinitely close to $t$ given $X$. Below let $r(t,x)$ denote the propensity scores. The right hand side of the equation represents the probability density function of a normal distribution. To estimate the propensity scores, we need to run a linear regression predicting the treatment by a set of covariates [@austin2019assessing]. From that we get the fitted values ($X\hat{\beta}$) and the model variance (${\sigma}^2$) [@austin2019assessing]. The fitted values take the place of the mean in the density function.

$$ r(t, x) = {f_{T|X}}^{(t|x)} = \frac{1}{\sqrt{2\pi\hat{\sigma}^2}} e^{-\frac{(t - X\hat{\beta})^2}{2\pi\hat{\sigma}^2}}$$

Conditional on the propensity scores, we can assume that each potential outcome is independent of treatment:

$$Y(t) \perp\!\!\!\perp T |r(t,x)$$

@hirano_imbens_2004 state that: "Within strata with the same value of $r(t,X)$, the probability that $T = t$ does not depend on the value of $X$.” I have seen $1$ and $I$ in front of the $(T = t)$, denoting the indicator function [@hirano_imbens_2004; @bia2008stata].

$$X \perp\!\!\!\perp 1(T = t)|r(t,x)$$

# Calculating Weights

Following the same logic as the inverse propensity weights (IPW) for the estimation of the average treatment effect (ATE) for a binary treatment, we calculate the inverse of the propensity scores as the weights:

$$\frac{1}{{f_{T|X}}^{(t|x)}}$$

However, @robins2000marginal noted that such weights can result in infinite variance [@austin2019assessing]. They suggested to use stabilized weights as follows:

$$\frac{{f_{T}}^{(t)}}{{f_{T|X}}^{(t|x)}}$$

Here the numerator represents the marginal density of treatment:

$${f_{T}}^{(t)} = \frac{1}{\sqrt{2\pi\hat{\sigma_t}^2}} e^{-\frac{(t - \mu_t)^2}{2\pi\hat{\sigma_t}^2}}$$

The stabilized weights make the distribution of the IPW narrower as there is less difference between the numerators and the denominators [@van2011ipw].

# Real Data Analysis Example

The data that I use here is from [High School Longitudinal Study of 2009](https://nces.ed.gov/surveys/hsls09/index.asp) to analyze the effect of dropping out of high school on later math achievement. The missing data in the original dataset have been replaced with one iteration of imputation using `mice` [@van2011mice]. This is not an appropriate method to analyze missing data but for the purpose of the example I am just using the one complete data. For the sake of this example, let's analyze the effect of math efficacy on later math achievement.

## Loading the Data

```{r, message = FALSE, warning = FALSE}
library(tidyverse)

dat <- read_csv("https://raw.githubusercontent.com/meghapsimatrix/datasets/master/causal/HSLS09_complete.csv")

```

## The Numerators

Here I am getting the numerators of the IPW, the marginal densities. I have regressed math_efficacy on just the intercept and used `dnorm` function to extract the densities.

```{r}
# the numerator
mod_num <- lm(math_efficacy ~ 1, data = dat)

num <- dnorm(x = dat$math_efficacy, # treatment 
             mean = fitted.values(mod_num), # fitted values
             sd = summary(mod_num)$sigma) # model sigma
```

## The Denominators

Here I am getting the denominators of the IPW, the conditional densities. I have regressed math_efficacy on $X$ and used `dnorm` function to extract the densities. I am not quite sure whether to use the model sigma which divides the sum of errors squared by the degrees of freedom before taking the square root or whether I should just take the standard deviation of the errors. However, with large sample size the difference between the two are negligible.

```{r}
# the demonimator
mod_den <- lm(math_efficacy ~ sex + race + language + repeated_grade + IEP + locale + region + SES, data = dat)

den <- dnorm(x = dat$math_efficacy, # treatment variable
             mean = fitted.values(mod_den), # fitted values
             sd = summary(mod_den)$sigma)

```

## The IPW

Below I calculate the stabilized weights:

```{r}
dat <- dat %>%
  mutate(ipw_s = num/den)

summary(dat$ipw_s)
```

## Checking Balance and Outcome Analysis

Short story: For balance, we have to calculate weighted correlations, and for outcome analysis we estimate the expected outcome for each treatment level and compare [@austin2019assessing].

# References
